---
title: "Assignment 1"
subtitle: "Data Visualization"
author: "Misha Ash"
date: "4/16/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(readr)
library(readxl)
library(data.table)
library(lubridate)
library(tidyverse)
library(purrr)
library(plyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
```

## Part 1: visualization critique

The [Allen Institute for Brain Science](https://www.alleninstitute.org/) has developed a range of publicly available research tools for neuroscientists, including a number of brain atlases. Among these is the [Reference Atlas of the Adult Human Brain](http://atlas.brain-map.org/), which is one of the most accurate representations of the structure of the human brain to date. According to [Ding et al. (2016)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5054943/):  

>[It is] the first digital human brain atlas to incorporate neuroimaging, high-resolution histology, and chemoarchitecture across a complete adult female brain, consisting of magnetic resonance imaging (MRI), diffusion-weighted imaging (DWI), and 1,356 large-format cellular resolution (1µm/pixel) Nissl and immunohistochemistry anatomical plates.

The team of scientists working at the Allen Institute collaborate with the global neuroscience community to develop and share information resources for advancing understanding of the brain. Moreover, the images that constitute the Human Brain Atlas were created through a highly controlled protocol. One possible limitation of the images might be that the slices are not equidistant (ranging from 0.4 mm to 3.4 mm), which is comparable to an axis with unequal units, though this may be relatively insignificant given the purposes of the images (it is also clearly stated). Therefore, this interactive visualization is truthful in its attempt to provide a more accurate, proportionate map of human brain structures.

The purpose of the this atlas is to inform and drive research, so it is designed to be useful primarily to scientists. The anatomical structures are highly detailed and can be individually selected. On the left pane, a hierarchical outline displays both abbreviations and full names of the structures, which puts each structure in the  context of the structural organization of the brain: "An essential component of modern interactive digital atlases is a unifying hierarchical structural ontology that provides unique IDs (and colors for representation) for each structure in a parent–child architecture" (Ding et al. 2010). The structure can thus be selected either by clicking it on the brain section or by name. The central image is a coronal slice of the left hemisphere, and each slice (from anterior to posterior) can be seen as a thumbnail along the bottom. It is entirely not clear why only the left lobe is displayed, particularly because some brain structures are known to be left or right lateralized (i.e., not fully symmetrical). A legend shows which slice of the lateral view is currently selected, which orients the viewer along the sagittal (front-to-back) axis, but its size varies with zooming making it difficult to see at some zoom levels. Interactivity also includes the capacity to zoom and select between alternative coordinate systems as background images. A legend shows scale (in microns). 

Furthermore, many elements can be toggled (description, metadata, thumbnails, outlines, ontology colors), selection color can be chosen, and individually configured atlas images can be downloaded. Therefore, despite its few apparent shortcomings are minor, and the features of this visualization make it both functional and valuable to researchers exploring the details of brain anatomy.

The aesthetics of the visualization are appealing and, combined with the interesting, complex, convoluted structure of the brain, it is beautiful. Despite the complex data presented, it does not appear cluttered and irrelevant details are not included. The tool can also be insightful, and indeed it is designed to foster insight. Highlighting structures and it is easy to imagine that seeing relations between structures at high resolution can yield insight into the architecture of the brain. One limitation of this tool might be the inability to navigate and explore these structures in three dimensions, however, which may be more insightful. 

Understanding the human brain---often recognized as the most complex piece of matter in the known universe---is no small feat, and every advance brings possibilities of developments of new treatments and interventions closer to the horizon. Advancing understanding of the brain is valuable and, in light of the prevalence and impact of mental illnesses, significant. The visualization may therefore also be said to be enlightening. 


## Part 2: ggplot2 and the grammar of graphics

```{r warning = FALSE, message = FALSE, echo = FALSE}
# transformed and anonymized data; removed original files
STAFF_PYRL <- read_csv(here::here("data", "STAFF_PYRL.csv"))
STAFF_CARE_C <- read_csv(here::here("data", "STAFF_CARE_C.csv"))

ggplot(STAFF_PYRL, aes(Date, Hours)) +

  geom_col(aes(fill = Name), color = "white", size = 0.28, alpha = .8) +
  scale_fill_brewer(type = "qual", palette = "Set3", direction = -1) +
  theme_minimal() +
  theme(#panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        legend.title = element_text(""), legend.position = "bottom" #,
        #axis.ticks.x = element_line()
        ) +

  labs(x = "", y = "Total Biweekly Hours", fill="", colour = "",
       title = "Office Staffing")
```

This visualization shows the hours of individual staff members color coded within each bar, and total staff hours worked within each biweekly period can be read off of the total height of each bar. The color scheme is chosen to provide contrast that is easy to distinguish without being excessively vivid, and the bars are divided with thin white lines to further enhance the segmentation. Although the graph displays time series data, for which lines are often preferable to clearly show change over time, in this case the bars enable a view of both total hours as well as the proportion of the hours accounted for by each staff member (via the color channel) for every period. When staff hours are instead shown with lines and broken down by each staff member the result is considerably more difficult to parse and a separate line for total staff hours would be needed:

```{r message = FALSE, warning = FALSE, echo = FALSE}
ggplot(STAFF_PYRL, aes(Date, Hours)) +

  geom_line(aes(Date, group=Name, color=Name), size = 0.38, alpha = .9) +
  #scale_color_brewer(type = "qual", palette = "Set3", direction = -1) +
  theme_minimal() +
  theme(#panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        legend.title = element_text(""), legend.position = "bottom") +

  labs(x = "", y = "Total Biweekly Hours", fill="", colour = "",
       title = "Office Staffing")

```

The bar chart is therefore a better solution for displaying both total staff hours and the breakdown of total hours by staff member.

Seeing the hours grouped by staff member is meaningful because it contributes to the purpose the graph serves, which is to contribute to informed executive decisions about whether staffing is adequate and how staffing needs are expected to change if current growth trends continue. Toward this end, two additional metrics are also displayed, this time with lines: 

* The relative amount of care provided for every hour staff work to oversee and manage the delivery of this care. 

* A projection of the number of staff hours will be needed in the coming months given the trend thus far. 

```{r warning = FALSE, message = FALSE, echo = FALSE}
ggplot(STAFF_PYRL, aes(Date, Hours)) +

  geom_col(aes(fill = Name), color = "white", size = 0.28, alpha = .7) +
  scale_fill_brewer(type = "qual", palette = "Set3", direction = -1) +
  theme_minimal() +
  theme(#panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        legend.title = element_text(""), legend.position = "bottom" #,
        #axis.ticks.x = element_line()
        ) +

  geom_point(data = STAFF_CARE_C, 
             aes(y = `CG PER STAFF`, color = "Care provided per staff hour worked"), size = .25) +
  
  geom_line(data = STAFF_CARE_C, 
              aes(Date, `CG PER STAFF`, color = "Care provided per staff hour worked"), 
              size = .45, alpha = .5) +
  
  stat_smooth(data = STAFF_CARE_C, 
              aes(Date, `STAFF HOURS`, color = "Projected staff hours"), 
              method = lm, fullrange = TRUE, size = .55, alpha = .45, fill = "lavender") +

  labs(x = "", y = "Total Biweekly Hours", fill="", colour = "",
       title = "Office Staffing",
       subtitle = "Staff Hours for Care Provided",
       caption = "SOURCE: COMPANY STAFF PAYROLL RECORDS") #+
  
  #guides(color = guide_legend(nrow = 1))
```

Given the narrower sizing of the bars, change information is apparent even without the use of lines for staff hours. Alpha values are adjusted to enhance contrast between the bars and lines. We can now see that in the third quarter of 2016 care provided for each staff hour worked begin to rise. Then, as new staff joined the office starting around mid-2017, this metric returned to levels comparable to those at the beginning of the timeline shown. The turquoise projection line extrapolates total staff hours with a linear model, providing a rough estimate of expected staff hours needed to maintain the average ratio of staff hours to caregiving service hours represented by the red line. 

Additional adjustments to the appearance of the visualization include moving the legends to the bottom, which provides more horizontal space for the graph, and using the minimal theme to provide a less cluttered appearance. Removing some of the grid lines was attempted but detracted from the ability to explore and make estimates, such as the approximate number of biweekly staff hours that may be needed by mid-2018 (325 ± 25). The data for this graph was taken directly from the company's payroll records.

